---
title: "evalGBM"
author: "J. Beaulieu"
date: "March 27, 2018"
output: pdf_document
classoption: 
  - landscape
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Results of evalGBM function for evaluation shrinkage rate and bag fraction values for brt model of data collected from 2016 reservoir survey.

```{r, echo=FALSE}
# volumetric
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalVolRateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalVolRateNat.RData")

# CH4 total
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalCh4trateFull.RData")

# CO2 diffusion
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalCo2drateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalCo2drateNat.RData")

# CO2 ebullition
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalCo2erateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalCo2erateNat.RData")


```

\pagebreak
**Volumetric ebullition rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalVolRateFull$plots[[1]]
evalVolRateFull$plots[[2]]
```

\pagebreak
**Volumetric ebullition rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalVolRateNat$plots[[1]]
evalVolRateNat$plots[[2]]
```


\pagebreak
**Total CH4 emission rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalCh4trateFull$plots[[1]]
evalCh4trateFull$plots[[2]]
```

\pagebreak
**CO2 diffusion rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalCo2drateFull$plots[[1]]
evalCo2drateFull$plots[[2]]
```

\pagebreak
**CO2 diffusion rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalCo2drateNat$plots[[1]]
evalCo2drateNat$plots[[2]]
```

\pagebreak
**CO2 ebullition rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalCo2erateFull$plots[[1]]
evalCo2erateFull$plots[[2]]
```

\pagebreak
**CO2 ebullition rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalCo2erateNat$plots[[1]]
evalCo2erateNat$plots[[2]]
```