---
title: "evalGBM"
author: "J. Beaulieu"
date: "Sept. 11, 2018"
output: pdf_document
classoption: 
  - landscape
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Results of evalGBM function for evaluation of shrinkage rate and bag fraction values for brt model of 2016 reservoir survey data.

```{r, echo=FALSE}
# volumetric
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalVolRateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalVolRateNat.RData")

# CH4 total
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCh4trateFull50.2856.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCh4trateNat.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalNatCh4trateNat.RData")

# CH4 diffusion
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCh4drateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCh4drateNat.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalNatCh4drateNat.RData")

# CH4 ebullition
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCh4erateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCh4erateNat.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalNatCh4erateNat.RData")

# CO2 diffusion
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCo2drateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCo2drateNat.RData")

# CO2 ebullition
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCo2erateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCo2erateNat.RData")

# CO2 total
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCo2trateFull.RData")
load("C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016/output/evalLocalCo2trateNat.RData")
```

\pagebreak
**Local volumetric ebullition rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalVolRateFull$plots[[1]]
evalLocalVolRateFull$plots[[2]]
```

\pagebreak
**Local volumetric ebullition rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalVolRateNat$plots[[1]]
evalLocalVolRateNat$plots[[2]]
```




\pagebreak
**Local total CH4 emission rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCh4trateFull$plots[[1]]
evalLocalCh4trateFull$plots[[2]]
```

\pagebreak
**Local total CH4 emission rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCh4trateNat$plots[[1]]
evalLocalCh4trateNat$plots[[2]]
```

\pagebreak
**National total CH4 emission rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalNatCh4trateNat$plots[[1]]
evalNatCh4trateNat$plots[[2]]
```





\pagebreak
**Local diffusive CH4 emission rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCh4drateFull$plots[[1]]
evalLocalCh4drateFull$plots[[2]]
```

\pagebreak
**Local diffusive CH4 emission rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCh4drateNat$plots[[1]]
evalLocalCh4drateNat$plots[[2]]
```

\pagebreak
**National diffusive CH4 emission rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalNatCh4drateNat$plots[[1]]
evalNatCh4drateNat$plots[[2]]
```






\pagebreak
**Local ebullitive CH4 emission rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCh4erateFull$plots[[1]]
evalLocalCh4erateFull$plots[[2]]
```

\pagebreak
**Local ebullitive CH4 emission rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCh4erateNat$plots[[1]]
evalLocalCh4erateNat$plots[[2]]
```

\pagebreak
**National ebullitive CH4 emission rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalNatCh4erateNat$plots[[1]]
evalNatCh4erateNat$plots[[2]]
```



\pagebreak
**Local CO2 diffusion rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCo2drateFull$plots[[1]]
evalLocalCo2drateFull$plots[[2]]
```

\pagebreak
**Local CO2 diffusion rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCo2drateNat$plots[[1]]
evalLocalCo2drateNat$plots[[2]]
```



\pagebreak
**Local CO2 ebullition rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCo2erateFull$plots[[1]]
evalLocalCo2erateFull$plots[[2]]
```

\pagebreak
**Local CO2 ebullition rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCo2erateNat$plots[[1]]
evalLocalCo2erateNat$plots[[2]]
```



\pagebreak
**Local CO2 total rate, all predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCo2trateFull$plots[[1]]
evalLocalCo2trateFull$plots[[2]]
```

\pagebreak
**Local CO2 total rate, national predictors**

```{r, echo=FALSE, out.width="50%"}
evalLocalCo2trateNat$plots[[1]]
evalLocalCo2trateNat$plots[[2]]
```