---
title: "Acton Lake Report"
author: "J. Beaulieu"
date: "January 22, 2017"
output: pdf_document
---

## Introduction

Introductory text desribing study and sampling specifics for the lake.

Fifteen samples sites were located on Acton Lake using a spatially-balanced
probabilistic survey design.


 2.  2D map of sampled sites.
 3.  3D scatterplot of 1) total CH4 emissions, 2) chl a, and 3) DO
 4.  Ordered dotplot of total CH4 emissions, TP, TN, and chl a.  
     Lake of interest should be highlighted with a different color.
 5.  Table of other water quality values at each site  
 6.  EPA disclaimer




```{r setup, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
#setwd('C:/Users/JBEAULIE/GitRepository/mulitResSurvey')
#knitr::opts_knit$set(root.dir = 'C:/Users/JBEAULIE/GitRepository/mulitResSurvey/ohio2016')
#myWd <- "C:/Users/JBEAULIE/GitRepository/mulitResSurvey"
globalWd <- "L:/Priv/Cin/NRMRL/ReservoirEbullitionStudy/multiResSurvey2016/grtsDesign/"

myWd <- "C:/R_Projects/mulitResSurvey"


```

Get the data and libraries
```{r results='hide', message=FALSE, warning=FALSE, echo=FALSE}
# knit() creates a new session and therefore can't read objects directly from
# workspace.  Run this file outside of project session and load project
# workspace.

load(paste(myWd, "/.RData", sep = ""))

# Libraries
library(ggplot2)
library(ggmap)
library(dplyr)
library(knitr)
library(rgdal)

# Load workspace from mulitResSurvey
load(paste(myWd, "/.RData", sep = ""))

```


```{r results='hide', message=FALSE, warning=FALSE, echo=FALSE, fig.cap='My Caption 1', fig.align='left', fig.height, fig.width}
## 2. 2D MAP OF SAMPLED SITES

# Read and project spatial points dataframe for plotting
actonSitesPlot <- readOGR(dsn = paste(globalWd, "acton", sep=""), 
                          layer = "actonSitesEqAreaData")  # shapefile with data

actonSites84 <- spTransform(x = actonSitesPlot, #reproject
                            CRS("+proj=longlat +datum=WGS84")) # projection for google maps

actonSites84@data <- mutate(actonSites84@data, 
                            long=coordinates(actonSites84)[,1], # add long to @data slot
                            lat=coordinates(actonSites84)[,2]) # add lat to @data slot

# Get ggmap
bbox <- make_bbox(data=actonSites84@data, #defines map extent based on sample site lat/lon
                  long, lat, f = 0.4) # f is zoom.  Large #, less zoom. tweak for each lake.  
actonSat <- get_map(location = bbox,
                    color = "color",
                    source = "google",
                    maptype = "satellite")

ggmap(actonSat) +
  ylab("Latitude") +
  xlab ("Longitude") +
  geom_point(data=filter(actonSites84@data, EvalStatus == "sampled"), #only main sites
             aes(x=long, y=lat),
             size = 2, color = "#00BFC4") + # specify color to be consistent across maps
  geom_text(data=filter(actonSites84@data, EvalStatus == "sampled"), #only main sites
            aes(label=siteID, x=long, y=lat),
            hjust=1.2, vjust=0, size=2, color = "#00BFC4") +      #alternate color: "#F8766D"
  coord_equal() +
  ggtitle("Acton Lake Sample Sites")
```

# 3.  3D scatterplot of 1) total CH4 emissions, 2) chl a, and 3) DO
```{r echo=FALSE, warning=FALSE, fig.width=7, fig.height=9}

actonEqArea <- readOGR(dsn = paste(globalWd, "acton", sep=""), # get polygon
                           layer = "actonEqArea")  # shapefile name

actonSitesPlot@data <- merge(actonSitesPlot@data, # add emission to point shp
                             filter(eqAreaData, Lake_Name == "Acton Lake") %>%
                               select(ch4.trate.mg.h, siteID))

# load external script (this is the scatterplot3d function, but changed a bit with to add more user control over style)
source(paste(myWd, "/ohio2016/scriptsAndRmd/scatterplot3d_edit.R", sep = ""))

# grab coordinates from polygon (in order to plot as a line on the xy plane of 3D plot)
coords = fortify(actonEqArea)
coords <- subset(coords, !hole)



##########################################################################
### Total CH4 emission rate
with(actonSitesPlot@data, {
  ch4_plot <- scatterplot3d_edit(x=coords$long, y=coords$lat, z=rep(0,length(coords$long)),   # x y and z data, respectively
                                   color="#282830",                               # color of lake outline
                                   lwd=2.0,                                       # line width of lake outline
                                   type="l",                                      # plot type - "l" = line, "h" = the lollipop style with a line  connecting to the horizontal plane 
                                   angle=40,                                      # angle aspect of plot (-40)
                                 scale.y=0.1,  
                                 #scale.y=0.5,                                   # scale y axis (increase by 175% (1.75))
                                   #main="Volumetric Gas Emissions at Harsha Lake",      # main plot title
                                   xlab="",                                       # x-axis label (left blank here to reposition later)
                                   ylab="",                                       # y-axis label
                                   cex.axis = 1.3,
                                   cex.lab = 2,
                                   mar = c(5,6,4,2)+0.1,
                                   zlab=expression(Volumetric~Ebullition~(mL~ m^{-2}~ day^{-1})),                      # z-axis label
                                   zlim=range(0,15),                            # range of z data shown - I grabbed the max from previously plotting the methane data here. Set so when I add the points, the z-axis fits that data. 
                                   lty.axis=1,                                    # the line type of the axes
                                   axis=TRUE,                                     # wheather to include axes
                                   
                                  grid=c('xy','xz','yz'),                        # which grids to show - added to edited version of scatterplot3d fx
                                   lty.grid.xy = 1,                               # grid line type  
                                   lty.grid.xz = 3,
                                   lty.grid.yz = 3,
                                   lwd.grid.xy = 1,                               # line width of grid 
                                   lwd.grid.xz = 1,
                                   lwd.grid.yz = 1,
                                   box=FALSE                                      # get rid of the box around the plot
  )
  

  # add the lollipop points
  ch4_plot$points3d(x=xcoord, y=ycoord, z=as.numeric(as.character(ch4.trate.mg.h)), # x y and z data
                      col="#282830",                      # color of lines connecting points to lake outline
                      lwd=1.5,                            # line width of lines connecting points to lake outline
                      pch=21,                             # type of point to plot
                      bg="#FF0000",                       # fill color of points
                      type="h",                           # plot type = lines to the horizontal plane
                      cex=2 #(vRateBest/500) + 0.5          # scaling of point sizes - this will need to be adjusted for each variable
  )
})
```

```{r echo=FALSE, warning=FALSE, fig.width=7, fig.height=9}

##########################################################################
### DO plot
with(actonSitesPlot@data, {
  do_plot <- scatterplot3d_edit(x=coords$long, y=coords$lat, z=rep(0,length(coords$long)),   # x y and z data, respectively
                                   color="#282830",                               # color of lake outline
                                   lwd=2.0,                                       # line width of lake outline
                                   type="l",                                      # plot type - "l" = line, "h" = the lollipop style with a line  connecting to the horizontal plane 
                                   angle=40,                                      # angle aspect of plot (-40)
                                   scale.y=0.5,                                   # scale y axis (increase by 175% (1.75))
                                   #main="Volumetric Gas Emissions at Harsha Lake",      # main plot title
                                   xlab="",                                       # x-axis label (left blank here to reposition later)
                                   ylab="",                                       # y-axis label
                                   cex.axis = 1.3,
                                   cex.lab = 2,
                                   mar = c(5,6,4,2)+0.1,
                                   zlab=expression(DO~(mg~ L^{-1})),                      # z-axis label
                                   zlim=range(0,20),                            # range of z data shown - I grabbed the max from previously plotting the methane data here. Set so when I add the points, the z-axis fits that data. 
                                   lty.axis=1,                                    # the line type of the axes
                                   axis=TRUE,                                     # wheather to include axes
                                   grid=c('xy','xz','yz'),                        # which grids to show - added to edited version of scatterplot3d fx
                                   lty.grid.xy = 1,                               # grid line type  
                                   lty.grid.xz = 3,
                                   lty.grid.yz = 3,
                                   lwd.grid.xy = 1,                               # line width of grid 
                                   lwd.grid.xz = 1,
                                   lwd.grid.yz = 1,
                                   box=FALSE                                      # get rid of the box around the plot
  )
  

  # add the lollipop points
  do_plot$points3d(x=xcoord, y=ycoord, z=as.numeric(as.character(DO__L_S)), # x y and z data
                      col="#282830",                      # color of lines connecting points to lake outline
                      lwd=1.5,                            # line width of lines connecting points to lake outline
                      pch=21,                             # type of point to plot
                      bg="#FF0000",                       # fill color of points
                      type="h",                           # plot type = lines to the horizontal plane
                      cex=2 #(vRateBest/500) + 0.5          # scaling of point sizes - this will need to be adjusted for each variable
  )
})



##########################################################################
### chlorophyll plot
with(actonSitesPlot@data, {
  chl_plot <- scatterplot3d_edit(x=coords$long, y=coords$lat, z=rep(0,length(coords$long)),   # x y and z data, respectively
                                   color="#282830",                               # color of lake outline
                                   lwd=2.0,                                       # line width of lake outline
                                   type="l",                                      # plot type - "l" = line, "h" = the lollipop style with a line  connecting to the horizontal plane 
                                   angle=40,                                      # angle aspect of plot (-40)
                                   scale.y=0.5,                                   # scale y axis (increase by 175% (1.75))
                                   #main="Volumetric Gas Emissions at Harsha Lake",      # main plot title
                                   xlab="",                                       # x-axis label (left blank here to reposition later)
                                   ylab="",                                       # y-axis label
                                   cex.axis = 1.3,
                                   cex.lab = 2,
                                   mar = c(5,6,4,2)+0.1,
                                   zlab=expression(DO~(mg~ L^{-1})),                      # z-axis label
                                   zlim=range(0, 45),                            # range of z data shown - I grabbed the max from previously plotting the methane data here. Set so when I add the points, the z-axis fits that data. 
                                   lty.axis=1,                                    # the line type of the axes
                                   axis=TRUE,                                     # wheather to include axes
                                   grid=c('xy','xz','yz'),                        # which grids to show - added to edited version of scatterplot3d fx
                                   lty.grid.xy = 1,                               # grid line type  
                                   lty.grid.xz = 3,
                                   lty.grid.yz = 3,
                                   lwd.grid.xy = 1,                               # line width of grid 
                                   lwd.grid.xz = 1,
                                   lwd.grid.yz = 1,
                                   box=FALSE                                      # get rid of the box around the plot
  )
  

  # add the lollipop points
  chl_plot$points3d(x=xcoord, y=ycoord, z=as.numeric(as.character(chla_S)), # x y and z data
                      col="#282830",                      # color of lines connecting points to lake outline
                      lwd=1.5,                            # line width of lines connecting points to lake outline
                      pch=21,                             # type of point to plot
                      bg="#FF0000",                       # fill color of points
                      type="h",                           # plot type = lines to the horizontal plane
                      cex=2 #(vRateBest/500) + 0.5          # scaling of point sizes - this will need to be adjusted for each variable
  )
})
```


## 4. DOT PLOTS

```{r echo=FALSE, warning=FALSE, fig.width=7, fig.height=3}
####################################################################################
### Item #4: Ordered dotplots of total CH4 emissions, TP, TN, and chla

# Highlight Harsha Lake (lake of interest) data with color
plotColor <- ifelse(meanVariance.c.lake.lu$Lake_Name == "William H Harsha Lake",
                    "red", "black")

# CH4 total emission rate
# Reset plotting order for CH4 total emissions
# "orderLake" is a function defined in the masterLibrary.R script
meanVariance.c.lake.lu$fLake_Name <- orderLake(meanVariance.c.lake.lu, 
                                               choice1 = "ch4.t")
ggplot(meanVariance.c.lake.lu,
       aes(ch4.trate.mg.h_Estimate, fLake_Name)) +
  geom_point(color = plotColor) +
  geom_errorbarh(aes(xmax = ch4.trate.mg.h_UCB95Pct, 
                     xmin = ch4.trate.mg.h_LCB95Pct), 
                 color = plotColor) +
  xlab(expression(CH[4]~emission~rate~(mg~ CH[4]~ m^{-2}~ hr^{-1}))) +
  theme(axis.title.y = element_blank()) +  # Eliminate x-axis title
  ggtitle(expression(Mean~CH[4]~emission~rate~
                       paste("(", "95% CI", ")", sep = "")))


# Chlorophyll 
meanVariance.c.lake.lu$fLake_Name <- orderLake(meanVariance.c.lake.lu, 
                                               choice1 = "chl")
ggplot(meanVariance.c.lake.lu,
       aes(chla_Estimate, fLake_Name)) +
  geom_point(color = plotColor) +
  geom_errorbarh(aes(xmax = chla_UCB95Pct, xmin = chla_LCB95Pct), color = plotColor) +
  xlab(expression(chl~a~paste("(", {mu}, "g ", L^{-1}, ")", sep = ""))) +
  theme(axis.title.y = element_blank())+
  ggtitle("Mean Chlorophyll a (95% CI)")


# TP 
# Reset plotting order to rank by TP
meanVariance.c.lake.lu$fLake_Name <- orderLake(meanVariance.c.lake.lu, choice1 = "TP")
ggplot(meanVariance.c.lake.lu,
       aes(tp_Estimate, fLake_Name)) +
  geom_point(color = plotColor) +
  geom_errorbarh(aes(xmax = tp_UCB95Pct, 
                     xmin = tp_LCB95Pct), 
                 color = plotColor) +
  xlab(expression(Total~Phosphorus~paste("(", {mu}, "g P ", L^{-1}, ")", sep = ""))) +   #NEED TO EDIT WITH UNITS
  theme(axis.title.y = element_blank()) +  # Eliminate x-axis title
  ggtitle("Mean TP (95% CI)")



# TN 
# Reset plotting order to rank by TN
meanVariance.c.lake.lu$fLake_Name <- orderLake(meanVariance.c.lake.lu, choice1 = "TN")
ggplot(meanVariance.c.lake.lu,
       aes(tn_Estimate, fLake_Name)) +
  geom_point(color = plotColor) +
  geom_errorbarh(aes(xmax = tn_UCB95Pct, 
                     xmin = tn_LCB95Pct), 
                 color = plotColor) +
  xlab(expression(Total~Nitrogen~paste("(", {mu}, "g N ", L^{-1}, ")", sep = ""))) +   #NEED TO EDIT WITH UNITS
  theme(axis.title.y = element_blank()) +  # Eliminate x-axis title
  ggtitle("Mean TN (95% CI)")




```

## Table

```{r warning=FALSE, echo=FALSE}
######################################################################
#### Item #5: Table of Water Quality Values by Site
## work in progress. Goal: produce a table for each lake with water quality and 
## GHG emissions info, including data that was not in the plots. Need to isolate
## the relevant data for each lake, rename the column headings, and add units

# dplyr package 
#"filter" command filters rows of interest. We're filtering the dataset for the 
# sampled sites at a given lake.
#"select" command selects the columns we want to show. The order they are listed 
#in the () is the order they will appear
actonTable <- filter(eqAreaData, Lake_Name == "Acton Lake", 
                     EvalStatus == "sampled") %>%
  select(siteID, LatSamp, LongSmp, wtrDpth,  #site identification  
         TNH4, TNO2.3, TN, TRP, TP,   #nutrient data
         chla_S, DO__L_S,  pH_S, SpCn__S, Tmp_C_S,  TrNTU_S) #shallow sonde

kable(actonTable,
      align = "c",
      digits = c(1,5,5,1,0,0,0,0,0,1,1,1,0,1,1),
      col.names = c("Site ID", "Latitude", "Longitude", "Depth (m)",
                    "NH4 (ugN/L)", "NO2.3 (ugN/L)", "Total N (ugN/L)", "Reactive P (ug/L)",
                    "Total P (ugP/L)", "Chl a (ug/L)", "DO (mg/L)", "pH",
                    "Sp. Cond", "Water Temp (C)", 
                    "Turbidity"))

```

